RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data

### 2014-03-31 doc/bioinformatics2014/document.tex
Update the manuscript

The remainder of this article is organized as follows. In the next section we describe the statistical model structure of our new algorithm, RVD2. Then, we derive a sampling algorithm for computing the posterior distribution over latent variables in the model and use those samples in a Bayesian posterior distribution hypothesis test to call variants. We compare the performance of RVD2 to several other variant calling algorithms for a range of read depths and minor allele fractions. Finally, we show that RVD2 is able to call variants on a heterogeneous clinical sample and identify two novel loss-of-heterozygosity events.
%TODO: Fixed this part
%%%%%%%%%%%%%%%%%%
% Model Structure
%%%%%%%%%%%%%%%%%%
\section{Model Structure}\label{sec:model_structure}

@@

\subsubsection*{Experimental process.}
Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control. Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele fractons (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%. More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.

\subsubsection*{Computational process.}
 We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads. To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96. The final data set contains read pairs for three replicates of each case and pairs of reads three replicates for the control sample giving $N=6$ replicates for the control and each MAF level.

\subsection{HCC1187 Sequence Data}
\subsubsection*{Experimental process.}
The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}. The HCC1187 cell line was derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. The estimated tumor purity was reported to be 0.8. Matched normal cells were derived from lymphoblastoid cells from peripheral blood. Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
\subsubsection*{Computational process.}
 The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads. We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
%Not sure the author is correct
% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx

>>>>

The remainder of this article is organized as follows. In the next section we describe the statistical model structure of our new algorithm, RVD2. Then, we derive a sampling algorithm for computing the posterior distribution over latent variables in the model and use those samples in a Bayesian posterior distribution hypothesis test to call variants. We compare the performance of RVD2 to several other variant calling algorithms for a range of read depths and minor allele fractions. Finally, we show that RVD2 is able to call variants on a heterogeneous clinical sample and identify two novel loss-of-heterozygosity events.

%%%%%%%%%%%%%%%%%%
% Model Structure
%%%%%%%%%%%%%%%%%%
\section{Model Structure}\label{sec:model_structure}

@@

\subsubsection*{Experimental Methods.}
Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control. Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele fractons (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%. More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.

\subsubsection*{Pre-processing Methods.}
 We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads. To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96. The final data set contains read pairs for three replicates of each case and pairs of reads three replicates for the control sample giving $N=6$ replicates for the control and each MAF level.

\subsection{HCC1187 Sequence Data}
\subsubsection*{Experimental Methods.}
The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}. The HCC1187 cell line was derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. The estimated tumor purity was reported to be 0.8. Matched normal cells were derived from lymphoblastoid cells from peripheral blood. Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
\subsubsection*{Pre-processing Methods.}
 The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads. We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
%Not sure the author is correct
% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx
